### 2016-12-05 chapter1.tex
Small modifications recommended by Makarevich

Closer to the Earth's surface, the magnetosphere also interacts with the Earth's ionosphere, a layer of partially ionized plasma in the upper atmosphere.  The interaction between charged particles in the ionosphere with magnetic field lines from the magnetosphere couples these two regions and creates non-trivial dynamics.  In addition, highly energized charged particles from the solar wind travel along open field lines and collide with both ions and neutral particles in the ionosphere, creating the aurora at high latitudes.  Below the ionosphere, the thermosphere consists mostly of neutral particles, however neutral winds can influence the ionosphere through viscous interactions along the lower boundary.  All of these regions contain incredibly complex dynamics and are additionally all interconnected, creating a complicated system that is extremely challenging to describe completely.

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The Earth's ionosphere is a region of the upper atmosphere that ranges approximately between 50--1000 km in altitude where neutral gases have been excited so that they are partially ionoized, resulting in free electrons and ions mixed with neutral particles.  Two conflicting processes occur in the ionosphere to create a peak in electron density, commonly referred to as the Chapman Layer \textbf{CTIE CHAPMAN?}.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increase as altitude decreases because there are more neturals available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination into neutral particles becomes a substantal factor, reducing the ion density.  The exact altitude where this density peak occurs is variable with time of day, location, season, and solar cycle.  A profile of how electron density changes with altitude is shown in Figure \ref{fig:densprofile}.

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption{Altitude profile of electron density throughout the ionosphere.  Typical ranges of the D, E, and F regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\subsubsection{SuperDARN}
\label{sec:superdarn}
The Super Dual Auroral Radar Network (SuperDARN) is a global network of HF CSRs that was designed to measure small-scale plasma structuring in the ionosphere and map the global plasma convection.  The network currently consists of about 35 operational radars between the northern and southern hemispheres distributed at mid-, high-, and polar latitudes, Figure \ref{fig:superdarnmap}.  Following the STARE experiment \citep{Greenwald1978}, the first HF ionospheric radar was built in Goose Bay, Canada in 1983, which would become the first of the SuperDARN radars \citep{Greenwald1985}.

\begin{figure}
	\includegraphics[width=\textwidth,angle=180]{superdarnmap.pdf}
	\caption{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

One of the original purposes of SuperDARN was to produce maps of the plasma convection patterns in the polar caps using 2D velocity vectors.  Originally this was accomplished by considering two radars with overlapping FoVs.  If both recieved backscatter from the same scattering volume, two LoS velocities could be found for that scattering volume, which could be combined to find a 2D velocity vector \citep{Ruohoniemi1989}.  The modern method for creating convection maps involves taking all data recorded by all radars in a particular hemisphere and finding the best fit to a 2D ionospheric electrostatic potential using Legendre functions \textbf{CITE???}.  For convection maps, only data from the F region is considered.  If there is not enough velocity data to confine the fit sufficiently, the measured data points are supplimented with a statistical model based on solar wind IMF conditions \textbf{CITE MCWILLIAMS???}.

@@

\section{A Brief Review of Plasma Structuring Theory and Observations}
Much work has been done over the past several decades to study the complicated picture of plasma structuring in the polar cap, including theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.  Here, a survey will be given of some of these advancements and the current state of understanding plasma structring on a range of scales.

\subsection{Theory and Simulations}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here we will focus on large regions of enhanced plasma density known as polar patches  \citep[][e.g.]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  The first observations of polar patches were made in the 1960s by \citet{Hill1963}.  They can range in size up to 1500 km in diameter and can have peak densities as much as eight times that of the background ionospheric plasma \citep{Weber1986,Hosokawa2016}.  Polar patches are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \citep{Weber1985,Weber1986}.

There have been at least 6 mechanisms proposed for how polar patches form on the dayside \citep{Crowley1996,Carlson2012}.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwook1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can ``cut'' segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{magnetosphere.pdf}
	\caption{Diagram of the interactions between the IMF, magnetosphere, and polar cap ionosphere.  Closed field lines in the Earth's magnetosphere are shown in grey while open field lines are light blue.  The polar cap open-closed boundary is in green.  Magnetic field lines that are part of the IMF are shown in dark blue.  Magnetic reconnection occurs along the X-line (red) on the dayside of the magnetopause (dashed line).  The X-line can be mapped back along magnetic field lines (orange) to the ionosphere, where their projection is the merging gap (pink dotted line).  Durring reconnection events, the open-closed boundary expands equatorwards as more open field lines are formed and there is plasma flow across the merging gap.}
	\label{fig:magnetosphere}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{patch_formation.pdf}
	\caption{Diagram of the polar patch formation through transient magnetopause reconnection.  The open-closed boundary is the green line.  Black contours represent plasma flow.  Highly ionized (dense) daytime plasma is shown by the light orange regions.  The burst of reconnection is occurring in Figures \ref{fig:patch_formation}a--\ref{fig:patch_formation}c but stops in Figures \ref{fig:patch_formation}d--\ref{fig:patch_formation}f.}
	\label{fig:patch_formation}
\end{figure}

@@

Recent advancements have improved the ability to image density structures in the ionosphere, particularly using multi-instrument approaches.  \citet{Semeter2009} introduced a method by which the density pattern in a particular volume could be imaged in three dimensions using AMISR systems.  This technique was later used by \citet{Dahlgren2012a,Dahlgren2012b} to image a polar patch.  \citet{Dahlgren2012b} additionally used ASIPs and SuperDARN radars to investigate the polar patch and found that backscatter tended to be observed more on the trailing edge of the patch.  This agrees well with many studies that have examined the asymmetry of small-scale plasma structuring surrounding plasma patches.  This asymmetry was first observed by \citet{Weber1984}.  Later, \citet{Milan2002b} found that backscatter power from HF radars tended to be greater on the trailing edges of moving structures, which was attributed to GDI being unstable on the trailing edge but stable on the leading edge.  \citet{Koustov2012} found similar results examining sun-aligned arcs.  HF radar backscatter power tended to be greater trailing behind a sun-aligned arc moving through the polar cap that was identified with ASIP data.  \citet{Moen2012} made direct measurements of plasma density structuring using the ICI-2 sounding rocket.  This study provided evidence that decameter scale plasma structuring (the same as observed by HF radars) had spawned from kilometer scale density gradients and structuring was greatest where GDI was operational.  \citet{Hosokawa2016} also considered the asymmetry of patches using an all-sky airglow imager (ASI) and found that the gradient on the leading edge of patches was typically 2--3 times steeper than that on the trailing edge.  In addition, they found large finger-like structures on the trailing edge of patches that were 10s--100s of kilometers in size.  These types of structures agree with GDI predictions, in particular they match the simulations of \citet{Gondarenko2004}.

@@

Scientifically, the ionosphere is an important part of the highly coupled Sun-Earth environment.  The polar cap ionosphere essentially serves as the boundary conditions for waves propagating through the magnetosphere.  In addition, understanding plasma structuring on a range of scales is important for correctly interpreting backscatter from ionospheric radars.  In particular, one of the main purposes of SuperDARN is to create large scale convection maps of both polar caps.  These convection maps have been used to give context of both magnetosphere and thermosphere behavior and are widely used in space physics research.  To map convection accurately, it is extremely important to understand when backscatter is observed and when it is absent, and why this is so.  Understanding plasma instability and wave growth mechanisms is very important for answering these kinds of questions.  Models and simulations of structuring in the polar caps incorporate structuring mechanisms so it is helpful if these are better understood, both the operation of a particular mechanism as well as how different mechanism are interconnected and which ones may be dominant under particular conditions.

The aim of this work is to examine many different factors behind plasma irregularity production in the F region polar cap using both observations and theory.  The majority of observational data has been found from ground based ionospheric radars, both CSR systems (Section \ref{sec:csr}) and ISR systems (Section \ref{sec:isr}), with particular emphasis on radars within the SuperDARN network, Section \ref{sec:superdarn}.  However, these datasets have also been supplemented with information from models as well as other instruments when necessary.  In addition, most experimental results are compared with predictions made by a linear fluid theory view of ionospheric plasma behavior.  Specific objectives are as follows.

@@

>>>>

Closer to the Earth's surface, the magnetosphere also interacts with the Earth's ionosphere, a layer of partially ionized plasma in the upper atmosphere.  The interaction between charged particles in the ionosphere with magnetic field lines from the magnetosphere couples these two regions and creates non-trivial dynamics.  In addition, highly energized charged particles from the solar wind travel along open field lines and collide with both ions and neutral particles in the ionosphere, creating the aurora at high latitudes.  Below the ionosphere, the thermosphere consists mostly of neutral particles, however neutral winds can influence the ionosphere through viscous interactions along the lower boundary.  Overall, the charged particle interactions in the ionosphere create a highly complex system, which is further complicated by coupling with both the thermosphere from below and the magnetosphere and solar wind from above.

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The Earth's ionosphere is a region of the upper atmosphere that ranges approximately between 50--1000 km in altitude where neutral gases have been excited so that they are partially ionoized, resulting in free electrons and ions mixed with neutral particles.  Two conflicting processes occur in the ionosphere to create a peak in electron density, commonly referred to as the Chapman Layer \textbf{CTIE CHAPMAN?}.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increase as altitude decreases because there are more neturals available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination into neutral particles becomes a substantal factor, reducing the ion density.  The exact altitude where this density peak occurs is variable with time of day, location, season, and solar cycle.  A profile of how electron density changes with altitude is shown in Figure \ref{fig:densprofile}, found from the International Reference Ionosphere (IRI) model run on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic (discussed further in Section \ref{sec:iri}).

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption{Altitude profile of electron density throughout the ionosphere found from the IRI model.  Typical ranges of the D, E, and F regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\subsubsection{Super Dual Auroral Radar Network (SuperDARN)}
\label{sec:superdarn}
The Super Dual Auroral Radar Network (SuperDARN) is a global network of HF CSRs that was designed to measure small-scale plasma structuring in the ionosphere and map the global plasma convection.  The network currently consists of about 35 operational radars between the northern and southern hemispheres distributed at mid-, high-, and polar latitudes, Figure \ref{fig:superdarnmap}.  Following the STARE experiment \citep{Greenwald1978}, the first HF ionospheric radar was built in Goose Bay, Canada in 1983, which would become the first of the SuperDARN radars \citep{Greenwald1985}.

\begin{figure}
	\includegraphics[width=1.3\textwidth,angle=90]{superdarnmap.pdf}
	\caption{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

One of the original purposes of SuperDARN was to produce maps of the plasma convection patterns in the polar caps using 2D velocity vectors.  Originally this was accomplished by considering two radars with overlapping FoVs.  If both recieved backscatter from the same scattering volume, two LoS velocities could be found for that scattering volume, which could be combined to find a 2D velocity vector \citep{Ruohoniemi1989}.  The modern method for creating convection maps involves taking all data recorded by all radars in a particular hemisphere and finding the best fit to a 2D ionospheric electrostatic potential using Legendre functions \citep{Ruohoniemi1998}.  For convection maps, only data from the F region is considered.  If there is not enough velocity data to confine the fit sufficiently, the measured data points are supplimented with a statistical model based on solar wind IMF conditions \citep{Ruohoniemi1998}.

@@

\section{A Brief Review of Plasma Structuring Theory and Observations in the Polar Cap}
Much work has been done over the past several decades to study the complicated picture of plasma structuring in the polar cap, including theoretical analysis of plasma dispersion relations, simulations of plasma dynamics, and observations of plasma structuring using a variety of instruments.  Here, a survey will be given of some of these advancements and the current state of understanding plasma structring on a range of scales.

\subsection{Theory and Simulations of Polar Patches and Plasma Instabilities}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here we will focus on large regions of enhanced plasma density known as polar patches  \citep[][e.g.]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  The first observations of polar patches were made in the 1960s by \citet{Hill1963}.  They can range in size up to 1500 km in diameter and can have peak densities as much as eight times that of the background ionospheric plasma \citep{Weber1986,Hosokawa2014}.  Polar patches are know to emerge on the dayside and then propagate with the background plasma convection across the polar cap to the nightside, where they either disintegrate or recombine with enhanced-density plasma in the auroral oval \citep{Weber1985,Weber1986}.

There have been at least 6 mechanisms proposed for how polar patches form on the dayside \citep{Crowley1996,Carlson2012}.
\begin{enumerate}
	\item Discrete changes in solar wind parameters, usch as IMF, density, speed, and pressure \citep{Sojka1994}
	\item Average flow patterns varying in time \citep{Anderson1988}
	\item Transient magnetopause reconnection \citep{Lockwood1992b}
	\item Plasma production by cusp particle precipitation \citep{Rodger1994,Millward1999}
	\item Plasma flow jet channels that cut continuous streaches of plasma into segments 	\citep{Valladares1998}
	\item Alfven wave coupling \citep{Prikryl1999}
\end{enumerate}
Although particle precipitation has been shown to create enhanced plasma density in the cusp region \citep{Roger1994}, it is unreasonable for this mechanism to account for the largest density enhancements observed in polar patches.  These high densities must originate from the resivor of solar illuminated plasma on the dayside.  Additionally, a mechanism is still required to separate patches from the cusp region.  High speed flow channels can ``cut'' segments off of a tongue of high-density plasma that has been pulled into the polar cap by convection \citep{Valladares1994,Valladares1998}.  Models have also shown changing large scale convection patterns quickly in time does produce density structures similar to patches \citep{Anderson1988}.  However, the mechanism that seems to be dominant for the majority of polar patches observed is transient magnetic reconnection \citep{Carlson2012}.

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{magnetosphere.pdf}
	\caption{Diagram of the interactions between the IMF, magnetosphere, and polar cap ionosphere.  Closed field lines in the Earth's magnetosphere are shown in grey while open field lines are light blue.  The polar cap open-closed boundary is in green.  Magnetic field lines that are part of the IMF are shown in dark blue.  Magnetic reconnection occurs along the X-line (red) on the dayside of the magnetopause (dashed line).  The X-line can be mapped back along magnetic field lines (orange) to the ionosphere, where their projection is the merging gap (pink dotted line).  Durring reconnection events, the open-closed boundary expands equatorwards as more open field lines are formed and there is plasma flow across the merging gap.}
	\label{fig:magnetosphere}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{patch_formation.pdf}
	\caption{Diagram of the polar patch formation through transient magnetopause reconnection.  The open-closed boundary is the green line.  Black contours represent plasma flow.  Highly ionized (dense) daytime plasma is shown by the light orange regions.  The burst of reconnection is occurring in Figures \ref{fig:patch_formation}a--\ref{fig:patch_formation}c but stops in Figures \ref{fig:patch_formation}d--\ref{fig:patch_formation}f.}
	\label{fig:patch_formation}
\end{figure}

@@

Recent advancements have improved the ability to image density structures in the ionosphere, particularly using multi-instrument approaches.  \citet{Semeter2009} introduced a method by which the density pattern in a particular volume could be imaged in three dimensions using AMISR systems.  This technique was later used by \citet{Dahlgren2012a,Dahlgren2012b} to image a polar patch.  \citet{Dahlgren2012b} additionally used ASIPs and SuperDARN radars to investigate the polar patch and found that backscatter tended to be observed more on the trailing edge of the patch.  This agrees well with many studies that have examined the asymmetry of small-scale plasma structuring surrounding plasma patches.

This asymmetry was first observed by \citet{Weber1984}.  Backscatter power from HF radars tends to be greater on the trailing edges of both moving high-density polar structures \citep{Milan2002b} and sun-aligned arcs identified from ASIP data \citep{Koustov2012}, which is generally attributed to GDI being unstable on the trailing edge but stable on the leading edge of such large, moving density structures.  Studies of patches using all-sky airglow imagers (ASI) found that the density gradient on the leading edge of polar patches tends to be 2--3 times steeper than that on the trailing edge \citep{Hosokawa2016}.  Additionally, large finger-like structures on the trailing edge of patches have been identified that were 10s--100s of kilometers in size and agree with predictions made by GDI simulations \citep{Gondarenko2004}.  \citet{Moen2012} made direct measurements of plasma density structuring using the ICI-2 sounding rocket.  This study provided evidence that decameter scale plasma structuring (the same as observed by HF radars) had spawned from kilometer scale density gradients (that observed by ASIP and ASI methods) and structuring was greatest where GDI was operational.

@@

The ionosphere is an important part of the highly coupled Sun-Earth environment.  The polar cap ionosphere essentially serves as the boundary conditions for waves propagating through the magnetosphere.  In addition, understanding plasma structuring on a range of scales is important for correctly interpreting backscatter from ionospheric radars.  In particular, one of the main purposes of SuperDARN is to create large scale convection maps of both polar caps.  These convection maps have been used to give context of both magnetosphere and thermosphere behavior and are widely used in space physics research.  To map convection accurately, it is extremely important to understand when backscatter is observed and when it is absent, and why this is so.  Understanding plasma instability and wave growth mechanisms is very important for answering these kinds of questions.  Models and simulations of structuring in the polar caps incorporate structuring mechanisms so it is helpful if these are better understood, both the operation of a particular mechanism as well as how different mechanism are interconnected and which ones may be dominant under particular conditions.

The aim of this work is to examine some of the factors (such as global solar control and GDI directional dependance) behind plasma irregularity production in the F region polar cap using both observations and theory.  The majority of observational data has been found from ground based ionospheric radars, both CSR systems (Section \ref{sec:csr}) and ISR systems (Section \ref{sec:isr}), with particular emphasis on radars within the SuperDARN network, Section \ref{sec:superdarn}.  However, these datasets have also been supplemented with information from models as well as other instruments when necessary.  In addition, most experimental results are compared with predictions made by a linear fluid theory view of ionospheric plasma behavior.  Specific objectives are as follows.

@@

The outline of this thesis is as follows.  Chapter \ref{sec:introduction} has provided an overview of the polar ionosphere and both macro- and microphysical plasma structuring mechanisms.  Chapter \ref{sec:paper1} will discuss plasma irregular occurrence and the extent to which solar factors control this occurrence.  Chapter \ref{sec:paper2} will consider the directional dependance of the GDI growth rate, and in particular give recommendations for the optimal orientation between a ground-based ionospheric radar and a large-scale density structure to observe small-scale structuring.  Chapter \ref{sec:paper3} will examine coincident observations made an IRS of background plasma density conditions and a CSR of small-scale plasma structuring in the polar cap and provide evidence for small-scale structuring being directionally dependent in a similar fashion to GDI predictions for large-scale plasma structuring.  Chapter \ref{sec:conclusion} will conclude this thesis and give a summary of the most significant results.  The implications of these results will be discussed along with questions that are still open and recommendations for future research.
